An Efficient Minibatch Acceptance Test for Metropolis-Hastings

### 2016-10-12 aistats2017.tex
Now about 8.3 pages

% Daniel: I'm moving this paragraph over here and putting the rest of it in the
% supplementary material since our code doesn't do this, so we should not
% describe this in much detail.
In fact we can do better than $O(n^{-1/2})$ error for the quantitative CLT
(showing the error decreases as $O(n^{-1})$) by using a more precise limit
distribution under an additional assumption. We review the details in
the Supplementary Material, Section~\ref{app:better_error_bound}, and leave the
implementation details to future work.

@@

We conduct two sets of experiments to explore the benefits of our minibatch MH
test and to benchmark it with previous work. In Section~\ref{ssec:gaussians}, we
show that our test enables samples to converge to the posterior distribution of
a heated Gaussian mixture model. In Section~\ref{ssec:logistic}, we analyze its
efficiency on logistic regression.

@@

\begin{figure*}[t]
    \centering
    \includegraphics[width=0.85\linewidth]{GaussianMixtureResult/posterior_of_gaussian.png}
    \caption{
    The log posterior contours and scatter plots of sampled $\theta$ values
    using different methods. 
    }
    \label{fig:gauss_mix_1}
\end{figure*}
\begin{figure*}[t]
    \centering
    \includegraphics[width=0.85\linewidth]{minibatch_size_gaussian.png}
    \caption{
    Minibatch sizes used in Section~\ref{ssec:gaussians}'s experiment. The axes
    have the same (log-log scale) range.
    }
    \label{fig:gauss_mix_2}
\end{figure*}

We borrow a Gaussian mixture experiment from~\citet{langevin_2011}.  The
parameter is 2-D, $\theta = \langle \theta_1,\theta_2 \rangle$, and the
generation process is
\begin{equation}\label{eq:data_generation}
\begin{split}
    \theta &\sim \mathcal{N}(0, {\rm diag}(\sigma_1^2,\sigma_2^2)) \\
    x_i & \sim 0.5 \cdot \mathcal{N}(\theta_1, \sigma_x^2) + 0.5 \cdot \mathcal{N}(\theta_1+\theta_2, \sigma_x^2).
\end{split}
\end{equation}
We set $\sigma_1^2 = 10, \sigma_2^2 = 1$ and $\sigma_x^2=2$.  We fix $\theta =
\langle 0,1 \rangle$. The original paper sampled 100 data points from this
distribution, and estimated the posterior in the parameters. We are interested
in performance on larger problems and so sampled 1,000,000 points to form the
posterior of $p(\theta)\prod_{i=1}^{1,000,000}p(x_i\mid \theta)$ with the same
prior from Equation~(\ref{eq:data_generation}). This produces a much sharper
posterior with two very narrow peaks.  Our goal is to reproduce the original
posterior, so we adjust the temperature to $T=10,000$.  Taking logs, we get the
target as shown in the far left of Figure~\ref{fig:gauss_mix_1}.

@@

For large $n$, the per-bin distributions are approximated by Poissons with
parameter $\lambda_i=P_i n$. Given samples $\{\theta_1,\ldots,\theta_T\}$, let
$c_j$ denote the number of individual samples $\theta_i$ that fall in bin $j$
out of $N_{\rm bins}$ total. We have
\begin{equation}\label{eq:log_prob_poissons}
\begin{split}
    \log p(c_1, \ldots, c_{N_{\rm bins}} \mid P_1, \ldots, P_{N_{\rm bins}}) & =  \\
    \sum_{j=1}^{N_{\rm bins}} c_j \log (n P_j) - n P_j - \log(\Gamma(c_j+1)). & \\
\end{split}
\end{equation}
The results for each model are shown in Table~\ref{tab:poissons}. It is
difficult, however, to interpret the scores. Thus, we perform significance tests
to show the difference between the MCMC-sampled distributions and the
ground-truth using the Chi-Squared distribution as the test statistic, which we
also show in Table~\ref{tab:poissons}. Both results imply that our method is
slightly superior to~\citet{cutting_mh_2014}, but slightly worse
than~\citet{icml2014c1_bardenet14}.

@@

@@

\begin{figure*}[t]
	\centering
	%\includegraphics[width=0.9\linewidth]{logistic_performance.png}
	\includegraphics[width=0.85\linewidth]{logistic_regression_ll_acc_results.png}
	\caption{
    Logistic regression performance (accuracy/log likelihood) based on
    cumulative data usage.
    }
	\label{fig:logistic_performance}
\end{figure*}
\begin{figure*}[t]
	\centering
	%\includegraphics[width=0.9\linewidth]{minibatch_size_logistic.png}
	\includegraphics[width=0.85\linewidth]{logistic_regression_histograms_results.png}
	\caption{
    Counts of minibatch sizes in the MNIST logistic regression experiment
    (analogous to Figure~\ref{fig:gauss_mix_2}).
    }
	\label{fig:logistic_minibatch}
\end{figure*}

@@

Figure~\ref{fig:logistic_performance} shows the testing log likelihood and
prediction accuracy, each as a function of the cumulative training data points
processed.\footnote{Note that the curves do not span the same length over the
x-axis, because our test consumes fewer samples throughout the MCMC procedure,
so the corresponding curve will ``end'' before the other two.} To generate the
curves, for each of the 3000 sampled vectors $\theta_t$,
$t\in\{1,\ldots,3000\}$, we use $\theta_t$ as the parameter for logistic
regression.  Our minibatch MH test is more efficient in terms of
accuracy, achieving convergence using roughly half as many elements
compared to~\citet{cutting_mh_2014} and \textbf{TODO TODO TODO} compared
to~\citet{icml2014c1_bardenet14}, though its log likelihood results lag slightly
behind the former algorithm during the very early stages.

@@

In this paper, we have derived a new MH test for minibatch MCMC methods. We
demonstrated how a simple deconvolution process allows us to use a minibatch
approximation to the full data tests. We experimentally show the benefits of our
test on Gaussian mixtures and logistic regression.  Straightforward directions
for future work include running more experiments with a particular focus on
investigation of the variance precondition.  More elaborate extensions include
testing on deep neural networks and combining our results with Hamiltonian Monte
Carlo methods, providing a recipe for how to use our algorithm (following the
framework of~\citet{sgmcmc_2015}), or integrating parallel
MCMC~\citep{conf/uai/AngelinoKWSA14,conf/icml/AhnSW14} concepts.

>>>>

% Daniel: I'm moving this paragraph over here and putting the rest of it in the
% supplementary material since our code doesn't do this, so we should not
% describe this in much detail.
In fact we can do better ($O(n^{-1})$) by using a more precise limit
distribution under an additional assumption. We review this in the Supplementary
Material, Section~\ref{app:better_error_bound}, and leave the details to future
work.

@@

We conduct two sets of experiments. In Section~\ref{ssec:gaussians}, we analyze
convergence to the posterior of a heated Gaussian mixture. In
Section~\ref{ssec:logistic}, we analyze its efficiency on logistic regression.

% Daniel: moving this here so figures don't overflow to next pages. We can tune
% the positioning if needed.
\begin{figure*}[t]
    \centering
    \includegraphics[width=0.85\linewidth]{GaussianMixtureResult/posterior_of_gaussian.png}
    \caption{
    The log posterior contours and scatter plots of sampled $\theta$ values
    using different methods. 
    }
    \label{fig:gauss_mix_1}
\end{figure*}
\begin{figure*}[t]
    \centering
    \includegraphics[width=0.85\linewidth]{minibatch_size_gaussian.png}
    \caption{
    Minibatch sizes used in Section~\ref{ssec:gaussians}'s experiment. The axes
    have the same (log-log scale) range.
    }
    \label{fig:gauss_mix_2}
\end{figure*}

@@

We borrow a Gaussian mixture experiment from~\citet{langevin_2011}.  The
parameter is 2-D, $\theta = \langle \theta_1,\theta_2 \rangle$, and the
generation process is
\begin{equation}\label{eq:data_generation}
\begin{split}
    \theta &\sim \mathcal{N}(0, {\rm diag}(\sigma_1^2,\sigma_2^2)) \\
    x_i & \sim 0.5 \cdot \mathcal{N}(\theta_1, \sigma_x^2) + 0.5 \cdot \mathcal{N}(\theta_1+\theta_2, \sigma_x^2).
\end{split}
\end{equation}
We set $\sigma_1^2 = 10, \sigma_2^2 = 1$ and $\sigma_x^2=2$.  We fix $\theta =
\langle 0,1 \rangle$. The original paper sampled 100 data points and estimated
the posterior. We are interested in performance on larger problems and so
sampled 1,000,000 points to form the posterior of
$p(\theta)\prod_{i=1}^{1,000,000}p(x_i\mid \theta)$ with the same prior from
Equation~(\ref{eq:data_generation}). This produces a much sharper posterior with
two very narrow peaks.  Our goal is to reproduce the original posterior, so we
adjust the temperature to $T=10,000$.  Taking logs, we get the target as shown
in the far left of Figure~\ref{fig:gauss_mix_1}.

@@

For large $n$, the per-bin distributions are approximated by Poissons with
parameter $\lambda_i=P_i n$. Given samples $\{\theta_1,\ldots,\theta_T\}$, let
$c_j$ denote the number of individual samples $\theta_i$ that fall in bin $j$
out of $N_{\rm bins}$ total. We have
\begin{equation}\label{eq:log_prob_poissons}
\begin{split}
    \log p(c_1, \ldots, c_{N_{\rm bins}} \mid P_1, \ldots, P_{N_{\rm bins}}) & =  \\
    \sum_{j=1}^{N_{\rm bins}} c_j \log (n P_j) - n P_j - \log(\Gamma(c_j+1)). & \\
\end{split}
\end{equation}
Table~\ref{tab:poissons} shows the results. It is difficult, however, to
interpret the scores, so we perform significance tests to show the difference
between the MCMC-sampled distributions and ground-truth using the Chi-Squared
distribution as the test statistic (also in Table~\ref{tab:poissons}). Both
imply that our method is slightly superior to~\citet{cutting_mh_2014}, but
slightly worse than~\citet{icml2014c1_bardenet14}.

@@

% Daniel: moving this here so figures don't overflow to next pages. We can tune
% the positioning if needed.
\begin{figure*}[t]
	\centering
	%\includegraphics[width=0.9\linewidth]{logistic_performance.png}
	\includegraphics[width=0.85\linewidth]{logistic_regression_ll_acc_results.png}
	\caption{
    Logistic regression performance (accuracy/log likelihood) based on
    cumulative data usage.
    }
	\label{fig:logistic_performance}
\end{figure*}
\begin{figure*}[t]
	\centering
	%\includegraphics[width=0.9\linewidth]{minibatch_size_logistic.png}
	\includegraphics[width=0.85\linewidth]{logistic_regression_histograms_results.png}
	\caption{
    Counts of minibatch sizes in the MNIST logistic regression experiment
    (analogous to Figure~\ref{fig:gauss_mix_2}).
    }
	\label{fig:logistic_minibatch}
\end{figure*}

@@

@@

Figure~\ref{fig:logistic_performance} shows the test log likelihood and
prediction accuracy as a function of the cumulative training points
processed.\footnote{Note that the curves do not span the same length over the
x-axis, because our test consumes fewer samples throughout the MCMC procedure,
so the corresponding curve will ``end'' before the other two.} To generate the
curves, for each of the sampled vectors $\theta_t$, $t\in\{1,\ldots,3000\}$, we
use $\theta_t$ as the parameter for logistic regression.  Our minibatch MH test
is more efficient in terms of accuracy, achieving convergence using roughly half
as many elements compared to~\citet{cutting_mh_2014} and \textbf{TODO TODO TODO}
compared to~\citet{icml2014c1_bardenet14}, though its log likelihood results lag
slightly behind the former algorithm during the very early stages.

@@

In this paper, we derived an M-H test for minibatch MCMC which approximates full
data tests. We provide theoretical results and experimentally show the benefits
of our test on Gaussian mixtures and logistic regression. Directions for future
work include running more experiments with a particular focus on controlling
variances.  More elaborate extensions include testing on neural networks,
combining our results with Hamiltonian Monte Carlo methods, providing a recipe
for how to use our algorithm (following the framework of~\citet{sgmcmc_2015}),
or integrating parallel MCMC~\citep{conf/uai/AngelinoKWSA14,conf/icml/AhnSW14}
concepts.
